Binary Latent Representations for Efficient Ranking

### 2017-06-13 paper/paper.tex
Address comments.

\usepackage{booktabs}
\usepackage{commath}

@@

In an on-line setting, where recommendation latency directly impacts the user experience, system designers have to work within a strict latency budget. Respecting that budget often means trading off model accuracy for speed, either via smaller and less expressive models (for example, with lower-dimensionality representations) or through aggressive use of heuristics to exclude large classes of candidate items from scoring. For example, approximate nearest neighbour search techniques are used at Youtube \citep{covington2016deep} to reduce the number of candidate videos that need to be scored for each user.

The challenge of low-latency scoring leads many systems to rely on a pre-compute system, where recommendations are computed in advance and stored for serving at a later time. The primary disadvantage of such a system is its inability to react dynamically to user actions (such as new interactions that allow the recommendations to be refined) and recommendation context (location, time of day or year). This leads to often complex architecture that mix fairly simple on-line models, computationally expensive offline models, and intermediate complexity near-line models where model updating is important \citep{amatriain2013big}. Additionaly, while a pre-compute solution removes model prediction from the critical path, it still requires substantial investment in compute and storage resources to calculate and cache the predictions of the model.

@@

\section{Related work}
\citet{shrivastava2014asymmetric} propose a solution for efficient ranking of large result sets through approximate maximum inner product search (MIPS). They extend well-known \citep{indyk1998approximate} locality-sensitive hashing techniques to the MIPS setting through asymmtric transformations of the query and candidate vectors, and show their applicability to recommender systems.

@@

In a traditional real-valued latent factor model, the prediction $r_{ui}$ for any user-item interaction pair $(u, i) \in \symUserSet \times \symItemSet$ is given by
\begin{equation}
r_{ui} = \vec{u}_i \cdot \vec{i}_i + b_i + b_u
\end{equation}
where $\vec{u}_i$ denotes the user and $\vec{i}_i$ the item $n$-dimensional latent vectors, and $b_u$ and $b_i$ the user and item biases.

\citet{rastegari2016xnor} show that the dot product of two real-valued vectors can be approximated in the binary domain by (1) binarizing the input vectors using the sign function (so that the results lie in $\{1, -1\}$), (2) taking their dot product, and (3) scaling the result by the average magnitude of the vectors' elements. Letting
\begin{equation}
\alpha = \frac{1}{n}\norm{\vec{i}_{i}}_{\ell1}
\end{equation}
and
\begin{equation}
\beta = \frac{1}{n}\norm{\vec{u}_{i}}_{\ell1}
\end{equation}
represent the scaling factors, the approximation is given by
\newcommand\binaryApproximation{ \alpha \beta\left(\sign(\vec{u}_i) \cdot \sign(\vec{i}_i)\right)}
\begin{equation}
\vec{u}_i \cdot \vec{i}_i \approx \binaryApproximation
\end{equation}
Applying this to the latent-factor model, the prediction for a user-item pair is given by
\begin{equation}
r_{ui} = \binaryApproximation + b_i + b_u
\end{equation}
Note that the scaling factors $\alpha$ and $\beta$ need to be stored (in addition to the latent representations and biases) to compute predictions in the binary model.

The model is trained using backpropagation. To facilitate the training of the binary model, we need a smooth approximation to the derivative of the sign function. Following \citet{courbariaux2016binarized}, we define
\begin{equation}
\frac{d\sign}{dw} = w1_{\abs{w} \leq 1}.
\end{equation}

@@

For the purposes of this paper, we focus on a simple bilinear collaborative filtering model, and do not use any external metadata information or model the sequential nature of the data. However, the binary dot produt approach generalizes beyond simple factorization models, and can be applied as a component of any model whose final scoring step involves a dot product between user and item representations. In particular, models using recurrent \citep{hidasi2015session} or convolutional \citep{lynch2015images} item or user representations can easily be augmented to use binary dot products in the final ranking stages. Assuming item representations can be computed in advance, even very complicated models can enjoy the efficiency advantages of using binary representations.

\section{Experiments}
\begin{table*}[htbp]
\caption{Movielens 100K results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio &   PPMS & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.096 &      0.088 &     0.919 & 89,299 &     245,644 &      2.751 &            0.091 \\
       64 & 0.098 &      0.096 &     0.973 & 47,565 &     209,486 &      4.404 &            0.062 \\
      128 & 0.108 &      0.091 &     0.843 & 20,502 &     156,404 &      7.629 &            0.047 \\
      256 & 0.109 &      0.097 &     0.884 & 10,631 &     166,500 &     15.661 &            0.039 \\
      512 & 0.112 &      0.105 &     0.938 &   5391 &     110,418 &     20.479 &            0.035 \\
     1024 & 0.114 &      0.102 &     0.897 &   2689 &      69,349 &     25.786 &            0.033 \\
     2048 & 0.111 &      0.109 &     0.984 &   1352 &      34,826 &     25.745 &            0.032 \\
     4096 & 0.111 &      0.107 &     0.961 &    666 &      15,907 &     23.856 &            0.032 \\
\bottomrule
\end{tabular}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 100K dataset \citep{harper2016movielens}. The dataset contains 100 thousand ratings from 943 users on 1682 movies. Despite its small size, the results on this dataset should be indicative of results that can be obtained on larger datasets, and its small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

@@

Results on scoring speed and memory use broadly confirm the naive calculations from section \ref{sec:approach}, converging to over 20 times faster scoring at around 3\% of memory as latent dimensionality increases.
As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss is around 8\%, varying between 2\% and 16\%.
Nevertheless, the binary model's speed enables some very favourable tradeoffs

>>>>

\usepackage{booktabs}
\usepackage{commath}
\usepackage{threeparttable}

@@

In an on-line setting, where recommendation latency directly impacts the user experience, system designers have to work within a strict latency budget. Respecting that budget often means trading off model accuracy for speed, either via smaller and less expressive models (with lower-dimensionality representations) or through aggressive use of heuristics to exclude large classes of candidate items from scoring. For example, approximate nearest neighbour search techniques are used at YouTube \citep{covington2016deep} to reduce the number of candidate videos that need to be scored for each user.

The challenge of low-latency scoring leads many systems to rely on a pre-compute system, where recommendations are calculated in advance and stored for serving at a later time. The primary disadvantage of such a system is its inability to react dynamically to user actions (such as new interactions that allow the recommendations to be refined) and recommendation context (location, time of day or year). This leads to often complex architectures that mix fairly simple on-line models, computationally expensive offline models, and intermediate complexity near-line algorithms where model updating is important \citep{amatriain2013big}. Additionaly, while a pre-compute solution removes model prediction from the critical path, it still requires substantial investment in computational and storage resources to calculate and cache the predictions of the model.

@@

\section{Related work}
\citet{shrivastava2014asymmetric} propose a solution for efficient ranking of large result sets through approximate maximum inner product search (MIPS). They extend well-known \citep{indyk1998approximate} locality-sensitive hashing techniques to the MIPS setting through asymmetric transformations of the query and candidate vectors, and show their applicability to recommender systems.

@@

In a traditional real-valued latent factor model, the prediction $r_{ui}$ for any user-item interaction pair $(u, i) \in \symUserSet \times \symItemSet$ is given by
\begin{equation}
r_{ui} = \vec{u}_u \cdot \vec{i}_i + b_u + b_i
\end{equation}
where $\vec{u}_u$ denotes the user and $\vec{i}_i$ the item $n$-dimensional latent vectors, and $b_u$ and $b_i$ the user and item biases.

\citet{rastegari2016xnor} show that the dot product of two real-valued vectors can be approximated in the binary domain by (1) binarizing the input vectors using the sign function (so that the results lie in $\{1, -1\}$), (2) taking their dot product, and (3) scaling the result by the average magnitude of the vectors' elements. Letting
\begin{equation}
\alpha = \frac{1}{n}\norm{\vec{i}_{i}}_{\ell1}
\end{equation}
and
\begin{equation}
\beta = \frac{1}{n}\norm{\vec{u}_{u}}_{\ell1}
\end{equation}
represent the scaling factors, the approximation is given by
\newcommand\binaryApproximation{ \alpha \beta\left(\sign(\vec{u}_u) \cdot \sign(\vec{i}_i)\right)}
\begin{equation}
\vec{u}_u \cdot \vec{i}_i \approx \binaryApproximation
\end{equation}
Applying this to the latent-factor model, the prediction for a user-item pair is given by
\begin{equation}
r_{ui} = \binaryApproximation + b_u + b_i
\end{equation}
Note that the scaling factors $\alpha$ and $\beta$ need to be stored (in addition to the latent representations and biases) to compute predictions in the binary model.

The model is trained using backpropagation, and so we a smooth approximation to the derivative of the sign function to facilitate training. Following \citet{courbariaux2016binarized}, we define
\begin{equation}
\frac{d\sign}{dw} = w1_{\abs{w} \leq 1}.
\end{equation}

@@

For the purposes of this paper, we focus on a simple bilinear collaborative filtering model, and do not use any external metadata information or model the sequential nature of the data. However, the binary dot product approach generalizes beyond simple factorization models, and can be applied as a component of any model whose final scoring step involves a dot product between user and item representations. In particular, models using recurrent \citep{hidasi2015session} or convolutional \citep{lynch2015images} item or user representations can easily be augmented to use binary dot products in the final ranking stages. Assuming item representations can be computed in advance, even very complicated models can enjoy the efficiency advantages of using binary representations.

\section{Experiments}
\begin{table*}[htbp]
\begin{threeparttable}
\caption{Movielens 100K results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.096 &      0.088 &     0.919 & 89,299 &     245,644 &      2.751 &            0.091 \\
       64 & 0.098 &      0.096 &     0.973 & 47,565 &     209,486 &      4.404 &            0.062 \\
      128 & 0.108 &      0.091 &     0.843 & 20,502 &     156,404 &      7.629 &            0.047 \\
      256 & 0.109 &      0.097 &     0.884 & 10,631 &     166,500 &     15.661 &            0.039 \\
      512 & 0.112 &      0.105 &     0.938 &   5391 &     110,418 &     20.479 &            0.035 \\
     1024 & 0.114 &      0.102 &     0.897 &   2689 &      69,349 &     25.786 &            0.033 \\
     2048 & 0.111 &      0.109 &     0.984 &   1352 &      34,826 &     25.745 &            0.032 \\
     4096 & 0.111 &      0.107 &     0.961 &    666 &      15,907 &     23.856 &            0.032 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond
}
\end{tablenotes}
\end{threeparttable}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 100K dataset \citep{harper2016movielens}. The dataset contains 100 thousand ratings from 943 users on 1682 movies. Despite its small size, the results on this dataset should be indicative of results that can be obtained on larger datasets, and its small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

@@

Results on scoring speed and memory use broadly confirm the naive calculations from section \ref{sec:approach}, converging to over 20 times faster scoring at around 3\% of memory as latent dimensionality increases.
As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss is around 8\%, varying between 2\% and 16\%.
Nevertheless, the model's speed enables some very favourable tradeoffs.
